Orienting the causal relationship between imprecisely measured traits using GWAS summary data

### 2017-07-18 docs/manuscript.Rmd
updated

tab_nums(
	"shakhtab", 
	caption="Summary of MR analysis on gene expression and DNA methylation associations. 'Count ($p_{Steiger} < 0.05$)' denotes the number of associations that have a p-value < 0.05 from the Steiger test. P(+ve effect) denotes the proportion of associations that were positive. $cor(\\rho_{MR}, \\rho_{P})$ is the correlation of the MR estimate obtained in this analysis and the Pearson correlation between reported in Shakhbazov et al 2016.", 
	display=FALSE)

@@

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal structure is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three basic competing models for these variables. The causal inference test (CIT) attempts to distinguish between them. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation. c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.",
	display=FALSE
)

@@

# fig_nums(
# 	"causality_exists_tpr",
# 	caption="Outcomes were simulated to be causally influenced by exposures with varying degrees of measurement imprecision applied to the exposure variable (x axis). True positive rates (y axis) for MR and CIT were compared for varying levels of measurement imprecision in the outcome variable (rows of boxes) and sample sizes (columns of boxes).",
# 	display=FALSE
# )

fig_nums(
	"bdmr", 
	caption="Using the same simulations as in Figure 2, but allowing an instrument for both the exposure and the outcome, bi-direcitonal MR was performed under different levels of measurement error in the exposure. The y-axis shows the probability for each of the three possible outcomes - obtaining the correct causal direction, obtaining the incorrect causal direction, or having insufficient power to detect a causal relationship. As measurement imprecision increases (decreasing values on x-axis) the power decreases but bi-directional MR is not liable to giving the wrong causal direction.",
	display=FALSE
)

fig_nums(
	"steiger_sensitivity_plot", 
	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface. When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by the range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction. Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y. A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and red surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$. In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$. b) Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative. When $d$ is negative the Steiger test is liable to infer the wrong direction of causality. Shaded regions show the parameter space where $d$ is negative. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.
	",
	display=FALSE
)

fig_nums(
	"cit_mr_comparison_figure", 
	caption="a) Outcome $y$ was simulated to be caused by exposure $x$ as shown in the graph, with varying degrees of measurement error applied to both. CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality. The value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality. Rows of graphs represent the sample size used in the simulations. For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality. b) As in (a) except the simulated model was non-causal, and a genetic confounder induces an association between $x$ and $y$. MR is unable to identify this model, so any significant associations are deemed to be incorrect. Outcome 3 denotes evidence for the correct model for the CIT method.",
	display=FALSE
)

fig_nums(
	"shakhplot",
	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them. a) The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels. The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes). If there is systematically higher measurement error in one platform than the other it will appear to be less likely to be the causal factor. b) The relationship between the Pearson correlation between DNA methylation and gene expression levels (x-axis) and the causal estimate (scaled to be in standard deviation units, y-axis). c) Distribution of estimated causal effect sizes, stratified into associations inferred to be due to DNA methylation causing expression (blue) and expression causing DNA methylation (red)."),
	display=FALSE
)

sfig_nums(
	"unmeasured",
	caption=paste0("Graph representing the unmeasured confounding parameters that will lead to the Steiger test returning the wrong causal direction. Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$)."),
	display=FALSE
)

@@

Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others. However, we show that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer. Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction between traits, with some advantages. First, it can be performed using only summary level data from genome-wide association studies; second, it is less susceptible to bias in the presence of measurement error or unmeasured confounding. We apply the method to infer the causal direction between DNA methylation and gene expression levels. Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms, and by horizontal pleiotropy. We emphasise that, where possible, implementing MR and appropriate sensitivity analyses alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.

@@

Understanding the causal relationships between pairs of traits is crucial for unravelling the causes of disease. To this end, results from genome-wide association studies are valuable because if a trait is known to be influenced by a genetic variant then this knowledge can be used to test the trait's causal influences on other traits and diseases. Here we demonstrate that when there is uncertainty about the nature of the genetic association several existing methods that are used for causal inference can be liable to give the wrong direction of causality. We introduce a new method that is more flexible and robust to problems such as measurement error, and apply it to evaluate the causal relationships between DNA methylation levels and gene expression. Our results show that DNA methylation is more likely to be the causal factor, but that systematic differences in measurement error between the platforms could influence such conclusions.

@@

Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004]. In response to this, the use of genetic associations to instrument traits has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015]. However, potential pitfalls remain and one that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.

@@

Measurement error will of course arise in other types of data too. For example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data have been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

@@

MR can be applied to the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner. Here the SNP is used as a surrogate for the exposure. Assuming the SNP associates with the outcome only through the exposure, the causal effect of the exposure on the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure. Though difficult to test empirically, this assumption can be relaxed in various ways when multiple instruments are available for a putative exposure [@Bowden2015; @Bowden2016b] and a number of sensitivity tests are now available to improve reliability [@Hemani2016].

@@

MR has some important advantages over the mediation-based approaches. First, the mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Second, the mediation-based approach of adjusting the outcome for the exposure to nullify the association between the SNP and the outcome is affected by unmeasured confounding of the exposure and outcome. This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and the in order to fully abrogate this association one must also adjust for all (hidden or otherwise) confounders. MR does not suffer from this problem because it does not test for association through adjustment. Third, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012]. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a metric to evaluate the sensitivity of the result of this extension to measurement error. Together these extensions improve the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this method to infer the direction of causation between DNA methylation levels and gene expression levels. Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].

@@

where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. The same model of measurement error can be applied to the outcome variable $y$. In this study we assume there is no measurement error in the SNP, and that measurement error in the exposure and the outcome are uncorrelated.

@@

- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then model 1 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then model 2 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the confounding model is accepted ($x \leftarrow g \rightarrow y$).

@@

To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement error in $x$ and $y$ we need to predict the values of $\rho_{gy} - \rho_{gx}$ for those values of measurement error. We offer two tools in which to do this. First, the user can provide values of measurement error for $x$ and $y$ and obtain a revised inference of the causal direction. Second, we integrate over the entire range of $\rho_{gy} - \rho_{gx}$ values for possible measurement error values, assuming that any measurement error value is equally likely. We find the ratio of the volume that agrees with the inferred direction of causality over the volume that disagrees with the inferred direction of causality. A ratio $R=1$ indicates that the inferred causal direction is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. In general, the $R$ value denotes that the inferred direction of causality is $R$ times more likely to be the empirical result than the opposite direction. Full details are provided in Appendix 2.

@@

where $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, $\alpha_{mx}$ and $\beta_{mx}$ are parameters that represent non-differential measurement error into the exposure variable $x$, and $\alpha_{my}$ and $\beta_{my}$ are parameters for non-differential measurement error in the outcome $y$. The simulations in `r cfig("bdmr")` are similar except we also simulate a second SNP which has a direct influence on $y$. MR can therefore be performed using the first SNP for estimating the effect of $x$ on $y$, and using the second SNP for estimating the effect of $y$ on $x$.

We have formulated the the non-causal model as:

$$
\begin{aligned}
g & \sim Binom(2, 0.5) \\
u & = \alpha_g + \beta_g g + \epsilon_g \\
x & = \alpha_u + \beta_u u + \epsilon_{u_x} \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_u + \beta_u u + \epsilon_{u_y} \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
$$

Here the $u$ variable mediates the influence of $g$ on $x$ and $g$ on $y$. The effects of $u$ on $x$ and $y$ are identical, therefore this simply attenuates the variance explained by $g$ on $x$ and $y$. All $\alpha$ values were set to 0, and $\beta$ values set to 1. Normally distribted values of $\epsilon_*$ were generated such that

$$
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(g, u)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
$$

@@

We performed simulations to verify that this problem does arise using the CIT method. `r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=1$) the CIT is reliable in identifying the correct causal direction. However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction. Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.

The bi-directional MR approach does offer a solution to this problem (`r cfig("bdmr")`)

@@

If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction. Here we introduce the MR Steiger approach to similarly orient the direction of causality but in an MR analysis when the underlying biology of the SNP is not fully understood.

For a particular association, it is of interest to identify the range of possible measurement error values agree and disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Appendix 2). This metric can be used to evaluate the reliability of MR Steiger.

We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2) determines the range of parameters around which the Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the Steiger test is likely to be correct about the causal direction). `r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$. It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the Steiger test to measurement error is relatively low.

@@

We used simulations to explore the performance of the MR Steiger approach in comparison to CIT in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect. Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.

>>>>

tab_nums(
	"shakhtab", 
	caption="Summary of MR analysis on gene expression and DNA methylation associations. 'Count ($p_{Steiger} < 0.05$)' denotes the number of associations that have a p-value < 0.05 from the MR Steiger test. P(+ve effect) denotes the proportion of associations that were positive. $cor(\\rho_{MR}, \\rho_{P})$ is the correlation of the MR estimate obtained in this analysis and the Pearson correlation between reported in Shakhbazov et al 2016.", 
	display=FALSE)

@@

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal structure is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three basic competing models for these variables. The causal inference test (CIT) attempts to distinguish between them. a) Gene expression causes methylation. The left figure shows that the SNP influences gene expression levels that in turn influence methylation levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from DNA methylation to Gene expression. c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.",
	display=FALSE
)

@@

fig_nums(
	"causality_exists_tpr",
	caption="Outcomes were simulated to be unrelated to the exposure (bottom plot) or causally influenced by the exposure (top plot) with varying degrees of measurement imprecision applied to the exposure variable (x axis). True positive rates (y axis) for MR and CIT were compared for varying sample sizes (columns of boxes).",
	display=FALSE
)

# fig_nums(
# 	"bdmr", 
# 	caption="Using the same simulations as in Figure 2, but allowing an instrument for both the exposure and the outcome, bi-direcitonal MR was performed under different levels of measurement error in the exposure. The y-axis shows the probability for each of the three possible outcomes - obtaining the correct causal direction, obtaining the incorrect causal direction, or having insufficient power to detect a causal relationship. As measurement imprecision increases (decreasing values on x-axis) the power decreases but bi-directional MR is not liable to giving the wrong causal direction.",
# 	display=FALSE
# )

fig_nums(
	"steiger_sensitivity_plot", 
	caption="a) We can predict the values the MR Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface. When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by the range of values where the blue surface is above the red plane, those values of measurement error lead to our observed MR Steiger test inferring the wrong causal direction. Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y. A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and red surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$. In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$. b) Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative. When $d$ is negative the MR Steiger test is liable to infer the wrong direction of causality. Shaded regions show the parameter space where $d$ is negative. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.
	",
	display=FALSE
)

fig_nums(
	"cit_mr_comparison_figure", 
	caption="a) Outcome $y$ was simulated to be caused by exposure $x$ as shown in the graph, with varying degrees of measurement error applied to both. CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality. The columns of graphs denote intervals for he value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the MR Steiger test to be more likely to be wrong about the direction of causality. Rows of graphs represent the sample size used in the simulations. For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality. b) As in (a) except the simulated model was non-causal, and a genetic confounder induces an association between $x$ and $y$. Neither CIT nor MR are able to identify this model, so any significant associations in MR are deemed to be incorrect, while outcomes 1 or 2 for the CIT are deemed to be incorrect.",
	display=FALSE
)

fig_nums(
	"shakhplot",
	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the MR Steiger test to infer the direction of causality between them. a) The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels. The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes). If there is systematically higher measurement error in one platform than the other it will appear to be less likely to be the causal factor. b) The relationship between the Pearson correlation between DNA methylation and gene expression levels (x-axis) and the causal estimate (scaled to be in standard deviation units, y-axis). c) Distribution of estimated causal effect sizes, stratified into associations inferred to be due to DNA methylation causing expression (blue) and expression causing DNA methylation (red)."),
	display=FALSE
)

sfig_nums(
	"unmeasured",
	caption=paste0("Graph representing the unmeasured confounding parameters that will lead to the MR Steiger test returning the wrong causal direction. Columns of boxes represent different signed values of the observational variance explained between $x$ and $y$ ($R^2_{xy}$)."),
	display=FALSE
)

@@

Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT). However, we show that measurement error in the phenotypes can lead to the CIT inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer. This problem is likely to be general to other mediation-based approaches. Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction between traits, with some advantages. First, it can be performed using only summary level data from genome-wide association studies; second, it is less susceptible to bias in the presence of measurement error or unmeasured confounding. We apply the method to infer the causal direction between DNA methylation and gene expression levels. Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms, and by horizontal pleiotropy. We emphasise that, where possible, implementing MR and appropriate sensitivity analyses alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.

@@

Understanding the causal relationships between pairs of traits is crucial for unravelling the causes of disease. To this end, results from genome-wide association studies are valuable because if a trait is known to be influenced by a genetic variant then this knowledge can be used to test the trait's causal influences on other traits and diseases. Here we demonstrate that when there is uncertainty about the nature of the genetic association existing methods that are used for causal inference can be liable to give the wrong direction of causality. We introduce a new method that is more flexible and robust to problems such as measurement error, and apply it to evaluate the causal relationships between DNA methylation levels and gene expression. Our results show that DNA methylation is more likely to be the causal factor, but that systematic differences in measurement error between the platforms could influence such conclusions.

@@

Observational measures of the human phenome are growing ever more abundant, but using these data to make causal inference is notoriously susceptible to many pitfalls, with basic regression-based techniques unable to distinguish a true causal association from reverse causation or confounding [@Phillips1991; @DaveySmith2002; @DaveySmith2004]. In response to this, the use of genetic associations to instrument traits has emerged as a technique for improving the reliability of causal inference in observational data, and with the coincident rise in genome-wide association studies it is now a prominent tool that is applied in several different guises [@DaveySmith2004; @Millstein2009; @Aten2008; @Waszak2015]. However, shifting from observational associations to instrumentation does require more (often untestable) assumptions, and potential pitfalls remain. One that is often neglected is the influence of non-differential measurement error on the reliability of causal inference.

@@

Measurement error will of course arise in other types of data too. For example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013], leading to the problem that phenotypes may be imprecisely defined. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data have been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

@@

MR can be applied to the same data - phenotypic measures of the exposure and the outcome variables and a genetic instrument for the exposure - but the genetic instrument is employed in a subtly different manner. Here the SNP is used as a surrogate for the exposure. Assuming the SNP associates with the outcome only through the exposure, the causal effect of the exposure on the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure. Though difficult to test empirically, this assumption can be relaxed in various ways when multiple instruments are available for a putative exposure [@Bowden2015; @Bowden2016b] and a number of sensitivity tests are now available to improve reliability [@Hemani2016]. Additionally, if valid genetic instruments are known for both traits of interest then MR can be performed in both directions (bi-directional MR), testing the influence of one trait on the other and vice versa, to infer the causal direction between the two phenotypes.

@@

MR has some important advantages over the mediation-based approaches. First, the mediation-based approaches require that the exposure, outcome and instrumental variables are all measured in the same data, whereas recent extensions to MR circumvent this requirement, allowing causal inference to be drawn when exposure variables and outcome variables are measured in different samples [@Pierce2013]. This has the crucial advantage of improving statistical power by allowing analysis in much larger sample sizes, and dramatically expands the breadth of possible phenotypic relationships that can be evaluated [@Hemani2016]. Second, the mediation-based approach of adjusting the outcome for the exposure to nullify the association between the SNP and the outcome is affected by unmeasured confounding of the exposure and outcome. This is because adjusting the outcome by the exposure induces a collider effect between the SNP and outcome [@Hernan2004], and in order to fully abrogate this association one must also adjust for all (hidden or otherwise) confounders. MR does not suffer from this problem because it does not test for association through adjustment. Third, when MR assumptions are satisfied the method is robust to there being measurement error in the exposure variable [@Pierce2012]. Indeed instrumental variable (IV) analysis was in part initially introduced as a correction for measurement error in the exposure [@Ashenfelter1994], whereas it has been noted that both classic mediation-based analyses [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013] and mediation-based methods that use instrumental variables [@Wang2015; @Lagani2015] are prone to be unreliable in its presence.

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. Though we only examine the CIT method in detail, we show that attempting to adjust for mediating variables to make causal inference is susceptible to problems, which can be generalised to other mediation-based methods. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a metric to evaluate the sensitivity of the result of this extension to measurement error. Finally, to demonstrate the potential impact of measurement error we apply this method to infer the direction of causation between DNA methylation levels and gene expression levels. Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].

@@

where $\alpha_{mx}$ and $\beta_{mx}$ influence the error in the measurement of $x$ by altering its scale, and $\epsilon_{mx}$ represents the imprecision (or noise) in the measurement of $x$. Measurement imprecision can represent imprecise measurement due to limits on sensitivity of measuring equipment, or due to phenotypes being imprecisely defined. The same model of measurement error can be applied to the outcome variable $y$.

In this study we assume there is no measurement error in the SNP. Common genetic variants are typically less susceptible to measurement error due to strict quality control procedures prior to genome wide association studies. Any non-differential measurement error that might be present (either because the SNP is poorly typed or because the SNP is not in complete linkage disequilibrium with the causal variant) will reduce power in MR but will not incur bias [@DaveySmith2004; @Pierce2012; @LeCessie2012]. We also assume that measurement error in the exposure and the outcome are uncorrelated.

@@

- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then model 1 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then model 2 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then there is likely confounding and no call is made ($x \leftarrow u \rightarrow y$).

@@

To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement error in $x$ and $y$ we need to predict the values of $\rho_{gy} - \rho_{gx}$ for those values of measurement error. We offer two tools in which to do this. First, the user can provide values of measurement error for $x$ and $y$ and obtain a revised inference of the causal direction. Second, we integrate over the entire range of $\rho_{gy} - \rho_{gx}$ values for possible measurement error values, assuming that any measurement error value is equally likely. Across all possible values of measurement error in $x$ and $y$ we find the volume that agrees with the inferred direction of causality and the volume that disagrees with the inferred direction of causality, and take the ratio of these two values. A ratio $R=1$ indicates that the inferred causal direction is highly sensitive to measurement error, because equal weight of the measurement error parameter space supporting both directions of causality. In general, the $R$ value denotes that the inferred direction of causality is $R$ times more likely to be the empirical result than the opposite direction. Full details are provided in Appendix 2.

@@

where non-differential measurement error is represented by a noise (measurement imprecision) term $\epsilon_{m*} \sim N(0, \sigma^2_{m*})$, and measurement bias terms $\alpha_{m*}$ and $\beta_{m*}$ for the exposure variable $x$ and the outcome variable $y$. Note that following the first section of the Results we no longer include the bias terms for simplicity. We have formulated the the non-causal model as:

$$
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_{gx} + \beta_{gx} g + \epsilon_{gx} \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_{gy} + \beta_{gy} g + \epsilon_{gy} \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
$$

All $\alpha$ values were set to 0, and $\beta$ values set to 1. Normally distributed values of $\epsilon_*$ were generated such that

$$
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
$$

@@

We performed simulations to verify that this problem does arise using the CIT method. `r cfig("cit_measurement_error_figure")` shows that when there is no measurement error in the exposure or outcome variables ($\rho_{x, x_o}=\rho_{y, y_o}1$) the CIT is reliable in identifying the correct causal direction. However, as measurement error increases in the exposure variable, eventually the CIT is more likely to infer a robust causal association in the wrong direction. Also of concern here is that increasing sample size does not solve the issue, indeed it only strengthens the apparent evidence for the incorrect inference.

@@

If we do not know whether the SNP $g$ has a primary influence on $x$ or $y$ then CIT can attempt to infer the causal direction. Though bi-directional MR can be used to orient causal directions, this requires knowledge of a valid instrument for each trait, and we were motivated to develop the MR Steiger method that could operate on summary data to orient the direction of causality using the same conditions as the CIT, where the underlying biology of a single SNP is not fully understood. We go on to explore the scenarios in which the method is likely to return the correct or incorrect causal directions.

We performed simulations to compare the power and type 1 error rates of MR and CIT in detecting a causal association between simulated variables under different levels of imprecision simulated in the exposure. Comparing the performance of methods with different sets of assumptions can be difficult, but a basic comparison is shown in `r cfig("causality_exists_tpr")`. We observe that the CIT is more conservative under the null model of no association owing to the omnibus test statistic comprising several statistical tests. The FDR using a p-value threshold of 0.05 appears to be close to 0, whereas for the MR Steiger method the FDR is around 0.05. Using the same p-value thresholds to declare significance in the non-null simulations, the general trend appears to be that the CIT power reduces as measurement error in the exposure increases faster than that of the MR Steiger method.

For a particular association, it is of interest to identify the range of possible measurement error values for which the method will give results that agree or disagree with the empirically inferred causal direction (`r cfig("steiger_sensitivity_plot")`a, Appendix 2). This metric can be used to evaluate the reliability of MR Steiger test.

We show that in the presence of measurement imprecision, $d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}$ (Appendix 2) determines the range of parameters around which the MR Steiger test is liable to provide the wrong direction of causality (*i.e.* if $d>0$ then the MR Steiger test is likely to be correct about the causal direction). `r cfig("steiger_sensitivity_plot")`b shows that when there is no measurement error in $x$, the MR Steiger test is unlikely to infer the wrong direction of causality even if there is measurement error in $y$. It also shows that in most cases where $x$ is measured with error, especially when the causal effect between $x$ and $y$ is not very large, the sensitivity of the MR Steiger test to measurement error is relatively low.

Unmeasured confounding between the exposure and outcome can also give rise to problems with the MR Steiger approach (Appendix 3). Though the relationship between unmeasured confounding and causal orientation is complex (`r sfig_nums("unmeasured")`), we note that when the magnitude of the observational variance explained between the exposure and the outcome is below 0.2 the MR Steiger method is unlikely to return the incorrect causal direction.

@@

We used simulations to explore the performance of the MR Steiger approach in comparison to CIT for different levels of measurement error. The performance was compared in terms of the rate at which evidence of a causal relationship is obtained for the correct direction of causality, and the rate at which evidence of a causal relationship is obtained where the reported direction of causality is incorrect. Simulations were performed for two models, one for a "causal model" where there was a causal relationship between $x$ and $y$; and one for a "non-causal model" where $x$ and $y$ were not causally related, but had a confounded association induced by the SNP $g$ influencing a confounder variable $u$.
